---
title: "Simmons et al. 2011"
author: "Rick Gilmore"
bibliography: [../include/bib/packages.bib, ../include/bib/psu-repro.bib]
csl: ../include/bib/apa.csl
---

## Purpose

This document simulates some data to illustrate Study 1 from [@simmons_false-positive_2011].
We discussed this study on [2023-10-04](../notes/wk07-2023-10-04.qmd) and [2023-10-06](../notes/wk07-2023-10-06.qmd).

## Generate data

There are two song groups:

```{r}
song <- c("kid", "ctrl")
song <- rep(song, each = 15)
song
```

There is one covariate, father's age.
Let's say that these are uniformly distributed between 40 and 70.

```{r}
father_age <- runif(n = 30, min = 40, max = 70)
hist(father_age)
```

And there is one dependent (outcome) measure that is the focus, “How old do you feel right now?”, with five options: very young, young, neither young nor old, old, and very old.
Let's give these ratings numbers between 1 and 5, and let's say that the average rating for the "kid" group was 3.75 and the average rating for the "ctrl" group was 3.
We'll assume that each of the groups ratings are normally distributed around these two means.

```{r}
old_feeling <- c(rnorm(15, mean = 3.75, sd = 0.5), rnorm(15, mean = 3, sd = 0.5))
hist(old_feeling)
```

But of course since respondents only give integer (1, 2, 3, 4, 5) ratings, we have to change the continuous scores to integer values.

```{r}
old_feeling <- round(old_feeling, 0)
hist(old_feeling, main = "Rounded `old_feeling`", breaks = c(1, 2, 3, 4, 5))
```

## Visualize data

To visualize the data, we'll make a data frame.
This is a convenient 'rectangular' format for our data.

```{r}
false_pos_study_1 <- data.frame(sub_id = 1:30, song, father_age, old_feeling)
str(false_pos_study_1)
```
Now, let's plot this to show how analysis of covariance (ANCOVA) might work.

:::  {.callout-note}

My graduate statistics professor, Dr. Mike Meyer, always exhorted us to "plot our data!"
That is, he thought we should always plot first, then run our statistics later.
I've taken that advice to heart in my own work.

:::

```{r fig-simmons-etal-2011-ancova, fig.cap="Illustration of a simulated ANCOVA based on [@simmons_false-positive_2011]"}
library(tidyverse)

false_pos_study_1 |>
  ggplot() +
  aes(x = father_age, y = old_feeling, color = song) +
  geom_point() +
  geom_smooth(method = lm) # linear fit
```

This plot adds two lines to the data, one for each song group.
The ANCOVA will let us evaluate whether the `old_feeling` is predicted by a linear combination of `father_age` and `song`.

In the code below, notice that we set `formula = old_feeling ~ song + father_age`. 
This is how we tell the `aov()` command that the linear combination of `song` and `father_age` are the variables we want to use to predict `old_feeling`.

```{r}
glm_ancova <- aov(formula = old_feeling ~ song + father_age, data = false_pos_study_1)
summary(glm_ancova)
```
Now, let's imagine that `father_age` wasn't entirely random for some reason.

```{r}
father_age <- (69:40) + rnorm(30, 0, 1)
plot(1:30, father_age)
hist(father_age)
```

```{r}
false_pos_study_1a <- data.frame(sub_id = 1:30, song, father_age, old_feeling)

false_pos_study_1a |>
  ggplot() +
  aes(x = father_age, y = old_feeling, color = song) +
  geom_point() +
  geom_smooth(method = lm) # linear fit

glm_ancova <- aov(formula = old_feeling ~ song + father_age, data = false_pos_study_1a)
summary(glm_ancova)
```

Or, maybe worse, that how old participants' felt related to their fathers' age.

```{r}
old_feeling <- (father_age-40)/(70-40) + rnorm(n=30, 0, 0.25)
old_feeling <- round(old_feeling/(max(old_feeling)-min(old_feeling))*4 + 1, 0)

false_pos_study_1b <- data.frame(sub_id = 1:30, song, father_age, old_feeling)

false_pos_study_1b |>
  ggplot() +
  aes(x = father_age, y = old_feeling, color = song) +
  geom_point() +
  geom_smooth(method = lm) # linear fit

glm_ancova <- aov(formula = old_feeling ~ song + father_age, data = false_pos_study_1b)
summary(glm_ancova)
```

