---
subtitle: "2024-09-20 Fri"
title: "Work session: Replication with R, R Markdown, & Quarto"
---

# Overview

## Announcement

- Retraction Watch [database](https://retractionwatch.com/retraction-watch-database-user-guide/) now [available to anyone](https://retractionwatch.com/2023/09/12/the-retraction-watch-database-becomes-completely-open-and-rw-becomes-far-more-sustainable/) via [CrossRef](https://www.crossref.org/blog/news-crossref-and-retraction-watch/)

## Last time...

- Wrap-up on replication attempt of [@Zhong2006-nf] by [@Earp2014-ek]
- Discuss replication attempt of [@Bargh1996-yv] by [@Doyen2012-ib]

## Today

- Wrap-up discussion of [@Bargh1996-yv] and [@Doyen2012-ib]; [Slides from Wed 9/13](wk04-2023-09-13.html)

---

*Work session: Replication with R, R Markdown, & Quarto*

- Review
  - [Survey 01 report](../surveys/survey-01.qmd)
  - [Survey 02 report](../surveys/survey-02.qmd)
- [Due]{.orange_due}
    - [Exercise 02: Textbook Findings](exercises/ex02-textbook-findings.qmd)
- [Class notes](wk04-2024-09-20-work-session-R.qmd)
  
# Replication with R, R Markdown, & Quarto

## Goals

::: {.incremental}
- Capture details of the research workflow
  - Not usually contained in journal articles, remember [@Artner2021-qg] comment discussed [earlier](../noteswk04-2024-09-16-replication-crisis.qmd)?
- Make research notebooks that actually do work
  - Not just capture info passively
- Do so efficiently, flexibly
:::

## More goals

::: {.incremental}
- Make *computer* work harder
- Reduce manual work
  - More error-prone
- Save history of changes (e.g., version control)
- Make use of HTML and the web
- Keep track of (& cite) resources
:::

## Research workflow

```{mermaid}
flowchart LR
  A{Idea/Question} -->|Develop|B{Plan}
  B -->|Create measures|C{Collect Info}
  C -->|Gather collected|D{Analyze Info}
  D -->|Visualize|E{Share findings}
  E -->|Evaluate|A
```

Graph made using [Mermaid.js](http://mermaid.js.org)

## Survey-01

- [Report](../surveys/survey-01.html)
- Goals
  - Measure trust in science and scientists
  - Measure individual characteristics that predict that trust
  - Do PSYCH 490.012 students follow patterns found by others [@Krumrei-Mancuso2016-qm; @Nadelson2014-ym; @Plohl2023-yf]?
  
## Survey questions

- Use [@Krumrei-Mancuso2016-qm; @Nadelson2014-ym; @Plohl2023-yf] questions.
- Shared [Google Form](https://forms.gle/471gpq6XXXq9iG9C8)

---

::: {.callout-important}
**Q**: Why Google Form?

**A**: Easier than Qualtrics to generate, extract information from. 

Also, more broadly available (Google Forms is free to people with a Google account).
:::
  
## Gather & Clean Data

- Use [Quarto](https://quarto.org); [@quarto-site-gw].
- Successor to RStudio's R Markdown

## {background-iframe="https://quarto.org"}

## Quarto project

- Set of text files that combine
  - Notes
  - Code (in the R programming language)
  - Web links
  - Figures
  
---

::: {.callout-important}
Q: Why not MS Word or Google Doc?

Q: Why not MS Excel or MS PowerPoint?
:::

---

```{mermaid}
flowchart TD
  A{Quarto File 1} --> B[HTML]
  C{Quarto File 2} --> B
  D{Quarto File 3} --> B
  A --> E(.docx)
  C --> E
  D --> E
  A --> F((.ppt))
  C --> F
  D --> F
```

## Anatomy of a Quarto [document](https://github.com/psu-psychology/psych-490-reproducibility-2024-fall/blob/main/src/surveys/survey-01.qmd)

````
---
title: "Survey 01"
subtitle: "Trust in science and scientists"
params:
  update_data: true
  fn: "survey-01-openness-trust-2024-fall.csv"
  make_plots: true
---
````

:::{.incremental}
- Document title & subtitle
- A parameter that determines whether the document should update the data or not.
- A parameter for the file name
- A parameter to control whether to make the plots or not.
:::

---

````
## Purpose

This page documents the data processing steps involved with Survey-01 in PSYCH 490.012 Fall 2024.

The survey questions were adapted from those discussed in [@Krumrei-Mancuso2016-qm; @Nadelson2014-ym; @Plohl2023-yf].
````

::: {.incremental}
- First section with header title ("Purpose")
    - *Note*: the double hash/pound symbols "##" mean this is a level 2 header
    - You do use MS Word's styles feature, don't you?
- Short-hand names for the references with the "@" symbol
    - Quarto generates in-text citations and a bibliography for us!
:::

---

````
Direct link: <https://forms.gle/kpwGFo2qpsKmNLya9>

<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSdvCvnemNhMyr0veX92ykxuCkHVrnqNWEGFXvbL6zYtggsepg/viewform?embedded=true" width="640" height="6027" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>
````

::: {.incremental}
- HTML links surrounded by "<>" become clickable in our output.
- the "<iframe..." comes from the Google Forms site. It embeds the actual survey in our document. 
    - No cutting and pasting!
:::

---

````
## Preparation

First, we load the external packages (groups of R commands) that we will be using.

```{{r}}
library('ggplot2')
library('dplyr')
```

````

::: {.incremental}
- Another level 2 header
- And some R code in a "code chunk".
  - The brackets "{r}" tell the computer that we're speaking R to it.
  - We could also speak other languages like "{python}", etc.
:::

## Enough anatomy

- [Gathering](../surveys/survey-01.html/#gathering) (downloading data from Google)
- Cleaning
- Visualizing
- Making a [data dictionary](https://en.wikipedia.org/wiki/Data_dictionary)

## Comparing this to other approaches...

## Let's look at [Survey-02](../surveys/survey-02.html)

![Summary figure from [Survey-02](../survey-02.html)](https://psu-psychology.github.io/psych-490-reproducibility-2023-fall/survey-02_files/figure-html/fig-boxplot-norm-counternorm-1.png){fig-align="center"}

## What other data picture(s) would you like to make?

## What analyses would you like to try?

## Evaluate predictions about mean levels across participants and items?

- Norms > Counternorms
- Norms: Should do > Actually do
- Counternorms: Should do < Actually do

## Prediction (table)

|              | Should do | Actually do |
|--------------|-----------|-------------|
| Norms        |   4       |      3.5    |
| Counternorms |   3       |      3.75   |

: Predictions made in-class on 2024-09-11 {#tbl-norm-preds}

## Prediction data picture (setup)

```{r, echo=TRUE}
#| label: set-up-barplot
library(tidyverse) # loads `tibble` and `ggplot2` packages
norm_df <- tibble(rating = c(4, 3.5, 3, 3.75), 
                          type = rep(c("norm", "counter"), each = 2),
                          resp_frame = rep(c("should", "actually"), times = 2))
norm_df
```

## Prediction data picture (plot)

```{r, echo=TRUE}
#| label: fig-kardash-2012-preds-barplot
#| fig-cap: "Data picture of [Survey 02](../surveys/survey-02.qmd) predictions in PSYCH 490.012 partial replication of [@Kardash2012-kq]"
norm_df |>
  ggplot() +
  aes(x = type, y = rating, fill = resp_frame) +
  geom_col() +
  facet_grid(cols = vars(resp_frame)) +
  xlab("") +
  ylim(NA, 5) +
  theme(legend.title = element_blank(), legend.position = "none")
```

## vs. Observed (table)

|              | Should do | Actually do |
|--------------|-----------|-------------|
| Norms        |   3.71       |      3.46  |
| Counternorms |   3.73       |      3.8   |

: [Survey-02](../surveys/survey-02.html#tbl-kardash-2012-all-items-observed) observations as of 2024-09-16.

## vs. Observed data picture (plot)

![Observed responses to [PSYCH 490.012 partial replication](../surveys/survey-02.qmd) of [@Kardash2012-kq]](../surveys/fig/fig-kardash-2012-all-items-observed-1.png)

## Did predictions == observations?

- Norms > Counternorms:
- Norms: Should do > Actually do: 
- Counternorms: Should do < Actually do:

## Did predictions == observations?

- Norms > Counternorms: **NO**
- Norms: Should do > Actually do: **YES**
- Counternorms: Should do < Actually do: **NO**

## Is this a better picture?

```{r, echo=TRUE}
#| label: fig-kardash-2012-preds-barplot-alt
#| fig-cap: "Data picture of [Survey 02](../surveys/survey-02.qmd) predictions in PSYCH 490.012 partial replication of [@Kardash2012-kq]"
norm_df |>
  ggplot() +
  aes(x = resp_frame, y = rating, fill = type) +
  geom_col() +
  facet_grid(cols = vars(type)) +
  xlab("") +
  ylim(NA, 5) +
  theme(legend.title = element_blank(), legend.position = "none")
```

# Next time

*Replication attempt: Priming effect*

- Read
    - [@Bargh1996-yv]; [PDF on Canvas](https://psu.instructure.com/courses/2350148/files/folder/readings?preview=165170715)
    - [@Doyen2012-ib]

# Resources

## References
